Comparing Hydrogen and Jet-A for an N+3 Turbofan with Water Recirculation using Gradient-Based Optimization

### 2023-05-06 conference-paper/main.tex
added revision edits and fixed plots

% ==================================================
%	Abstract
% ==================================================
\begin{abstract}
    Advances in commercial propulsion technology led to the development of efficient high bypass ratio turbofan engines with larger overall pressure ratios and internal temperatures.
    Current trends suggest that geared ultra high bypass ratio turbofans are the next generation of commercial propulsion systems.
    Furthermore, the emphasis on decreasing emissions has driven the exploration of hydrogen-powered aircraft, adding to the already challenging design space.
    Carrying and burning hydrogen introduces complexity and weight penalties that we must offset using the fuel's thermodynamic and chemical properties.
    In this study, we create a closed-loop water recirculation system with a zero-dimensional thermodynamic model and compare the benefits between jet-A and hydrogen fuels.
    We perform a gradient-based optimization parameter sweep to explore the trade-offs between performance and emissions using both fuels with water recirculation.
    The results quantify the design space for next-generation propulsion concepts that can take advantage of hydrogen fuel's advantageous thermodynamic properties to reduce emissions and improve performance.
\end{abstract}

\section{Introduction}
% Message 1: Motivation to incorporate low-emission fuels and techniques
The effects of climate change are pushing the aviation industry towards hydrogen-fueled propulsion systems as a solution to reduce emissions.
N+3 technology estimates for engines that burn hydrocarbon fuels suggest that higher efficiency can be achieved by designing ultra high bypass ratio (UHBR) turbofans with small cores and high overall pressure ratios (OPR).
Higher OPR and smaller cores challenge the limits of compressor and turbine design, placing an upper bound on potential performance and emissions improvements.
Switching to hydrogen as the primary fuel source reduces carbon dioxide emissions immediately, but adds complexity and weight that offset the benefits~\cite{Adler2023}.
However, hydrogen is a versatile fuel with advantageous molecular and thermodynamic properties that can be exploited to increase the performance and reduce emissions.
We introduce a closed-loop water recirculation model that demonstrates the possible efficiency gain when hydrogen is used for purposes other than combustion.
% [x] TODO: PA-PA, add section about MAUD and cite.

% Message 2: Background and references to support using H2 and water injection in HBTF engines
% [x] TODO: AL-PA, Need to find a journal/conference paper to cite for the following statement.  Maybe find one or more papers other than just the NASA report?
Water injection is the process of introducing water upstream of the combustor as finely atomized droplets.
NASA, Boeing, and Rolls-Royce studied this concept and suggested that this technique reduces the NOx emissions as much as 47 percent~\cite{Daggett2010}.
Additionally, water injection improves fuel efficiency and thrust output with lower combustion temperatures that can improve the lifetime of turbine blades and reduce noise~\cite{Daggett2010}.
% [x] TODO: AL-PA, Should try to find a citation for this if one exists.
Traditional propulsion systems that burn hydrocarbon fuels would require external water storage on the aircraft for water injection~\cite{Mourouzidis2015}.
The added weight of tanks, pumping, and ducting makes this concept infeasible for a conventional aircraft.
The main product of hydrogen combustion is water vapor that we can recover from the exhaust stream~\cite{Strom2002}.
Condensing water vapor from the exhaust stream of hydrogen combustion and recirculating it eliminates any additional storage requirements.
This allows for the theoretical design of a closed loop water feedback system inside the propulsion cycle.
Pratt and Whitney are actively researching this technology to improve the feasibility of hydrogen-powered propulsion~\cite{arpa-e_2021}.
They suggest that water vapor can be recovered by utilizing the thermal properties of hydrogen to condense some of the water in the turbofan exhaust.

@@

% Message 3: Introduce the extension of the HBTF and propose novel contributions
In this work, we analyze the thermodynamic benefits of a closed-loop water vapor recovery and water injection system in a high-bypass turbofan engine.
We develop pyCycle components for water injection and vapor recovery to quantify the benefit of a closed loop recirculation system.
We use gradient-based optimization to minimize fuel burn subject to performance requirements using both jet-A and hydrogen at a range of flight conditions.
The optimized results show the trade-off between complexity, performance, and efficiency for jet-A and hydrogen fuels.

@@

% Engine Architecture: Describe the flow path of the engine and establish the mechanical coupling.
The UHB turbofan model is the NASA advanced technology "N+3" engine~\cite{Jones2017a}.
The N+3 reference cycle is a UHB ratio geared turbofan that could be available in the 2030 to 2040 time frame.
The flow path consists of an inlet that directs ambient air through a fan, followed by a duct that splits the flow into a core and a bypass stream, each ending in a core and bypass nozzle, respectively.
The low pressure system is split into two mechanical subsystems.
First, the fan is connected to the gearbox that reduces the shaft speed to decrease fan tip speeds.
Second, the gearbox attaches to the low-pressure shaft that connects to the low-pressure compressor (LPC) and low-pressure turbine (LPT).
The high pressure compressor (HPC) is connected to the high pressure turbine (HPT) by the high pressure shaft.
% [x] TODO: AL-PA, I think we need to reference the engine diagram figure here so the reader can visualize all this text.
Figure~\ref{fig:n3_cycle} shows a standard high bypass turbofan engine flow connections.
% [x] TODO: AL-PA, potentially move the commented line below about CEA into the subsection discussing the propulsion model.  I think it's too much detail for the setup of this section.
We introduce the closed-loop water recovery system as a feedback cycle that transports water from the exhaust to upstream of the compressors.
The recovery system injects vaporized water into the core stream that reduces the combustion temperature due to heat absorption.
The vapor recovery component is placed directly before the core nozzle to extract water from the exhaust and recycle it back to the injector.
Figure~\ref{fig:n3_clvr} shows the water vapor recovery loop implemented in a turbofan engine.
% [x] TODO: AL-PA, Again, probably should reference a figure to give context to the water recovery layout.
% NOTE: AL-PA, don't discredit your work before you even explain the model and methods.  We can make this point in the results or conclusion.

@@

% [x] TODO: AL-PA, Remove the black box with the component descriptions from the bottom left of the figure.
% [x] TODO: AL-PA, Add the description of the figure elements to the caption.
% [x] TODO: AL-PA, Consider changing the caption to "Simplified layout of the N+3 engine cycle, adapted from~\citet{Hendricks2019}"
% [x] TODO: AL-PA, Remove the flight conditions component and add the bypass duct component below the splitter. I think the bypass is an important enough aspect of the engine to include.
\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_cycle.pdf}
    \caption{
        Simplified layout of the N+3 engine cycle, adapted from~\citet{Hendricks2019}.
        Black arrows are flow connections, red lines are mechanical connections, blue boxes are cycle elements, red boxes are shaft elements.
        Flow stations are identified with bubble text
    }
    \label{fig:N3_original}
\end{figure}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_CLVR_cycle.pdf}
    \caption{
        Simplified layout of the N+3 engine cycle with the closed-loop vapor recovery.
        Black arrows are flow connections, blue boxes are cycle elements
        Flow stations are identified with bubble text.
    }
    \label{fig:n3_clvr}
\end{figure}

@@

The ``on-design'' point is top-of-climb (TOC) with ``off-design'' conditions at rolling takeoff (RTO), sea-level static (SLS), and cruise (CRZ).
To account for each of these operating conditions, the N+3 model uses a technique called Multipoint Design Point (MDP) modeling to converge the model to an engine design.
Table~\ref{tab:flight_conds} shows the different altitudes and mach numbers for each flight condition.
We consider these points because they either limit the performance at cruise or present critical design considerations for the propulsion system.
% [x] TODO: AL-PA, Add a small table to show the mach and altitude of each flight condition
\begin{table}[hbt!]
    \centering
    \caption{Altitude and mach numbers at each of the flight conditions considered in the multipoint formulation.
        The humidity ratios were computed from humidity data~\cite{Kalnay1996}.
    }
    \begin{tabular}{l r r r r l}
        \hline
        Parameter      & TOC     & RTO   & SLS   & CRZ     & Units      \\
        \hline
        Altitude       & 35000   & 0     & 0     & 35000   & \si{ft}    \\
        Mach           & 0.8     & 0.25  & 0.0   & 0.8     & -          \\
        Humidity Ratio & 0.00017 & 0.009 & 0.009 & 0.00017 & \si{kg/kg} \\
        \hline
    \end{tabular}
    \label{tab:flight_conds}
\end{table}
% [x] TODO: AL-PA, You should explain why each design point is important in a separate sentence for each.
%   - For example, the SLS point is important for making sure the turbo machinery and flow paths are designed to meet a static thrust target
%   - The RTO point is limiting for temperature, etc...
% [x] TODO: AL-PA, Consider re-writing this section using the comments from the previous TODO.  Be more concise and direct with the language here.
% [x] TODO: AL-PA, This is a big run-on sentence.  Consider breaking it up and being more specific about the requirements at each design point.  See previous comments.
The design rules at SLS ensure that the turbo machinery and flow paths meet the static thrust target.
Rolling takeoff limitations constrain the upper limit of the combustor temperature and subsequently the cooling requirements of the turbines.
The cooling mass flow rates ($\dot{m}_{\rm{cool},k}$) are passed from the RTO point to all other flight conditions creating a cyclic connection.
We add a fuel burn objective at cruise to optimize the engine performance during the longest flight segment.
We size the water recovery system at the CRZ condition because it has the greatest impact on fuel burn over the flight envelope.
Similarly to $\dot{m}_{\rm{cool},k}$, the water recovery mass flow rates ($\dot{m}_{\rm{water},k}$) are connected back to the other operating points.
The interconnections between flight conditions create a complex nonlinear loop, shown in Figure~\ref{fig:N3_xdsm_full}.
The converged model represents a feasible multipoint engine that accounts for design considerations at all four operating conditions.
% [x] TODO: AL-PA, I removed a lot of stuff here to simplify this section.
%                  - Ultimately I think we need to reference the pyCycle paper to prevent duplicating a bunch of math/text
%                  - If you are okay with this check off this TODO.  If not feel free to add a bit more and then check it off.
We direct the reader to the N+3 multipoint model definition in the paper by \citet{Hendricks2019} for the detailed description of the balance relationships.

@@

Typical gas-turbine analysis measures the efficiency using the thrust-specific fuel consumption (TSFC).
% For measuring the efficiency of jet engines, thrust-specific fuel consumption (TSFC) is used in cycle analysis since it represents how much fuel is burned at a given thrust level.
However, this metric is not useful in the comparison of efficiency between jet-A and hydrogen fueled cycle models because of the differences in fuel density and energy capacity.
% However, when comparing Jet-A and hydrogen fuels this is not such a good metric since a given mass flow rate of hydrogen has an energy content almost 3 times that of Jet-A.
A better metric for comparing Jet-A versus hydrogen is thrust-specific energy consumption (TSEC)~\cite{Adler2023}.
TSEC is TSFC multiplied by the lower heating value (LHV) of the fuel shown in Equation~\eqref{eq:tsec}.

@@

For this work we are concerned with minimizing fuel burn of the engine subject to known thrust constraints.
% [x] TODO: AL-PA, TSEC is not the objective.  Fuel burn is the objective.  TSEC is the efficiency metric of interest.
TSEC will be used as a efficiency metric at the cruise condition to compare the two fuels in the results.
% [x] TODO: AL-PA, explain the important aspects of the XDSM in writing before referencing the figure.
Large multidisciplinary optimization problems are commonly represented with extended design structure matrix (XDSM) diagrams.
XDSM diagrams are used to visualize the complex implicit and explicit relationships between functions, the solver, and optimizer.
An XDSM diagram of the optimization problem with the multipoint formulation is shown in Figure~\ref{fig:N3_opt_xdsm}.

@@

% [x] TODO: AL-PA, I would break this up into two or more sentences.
The design variables for this problem are pressure controls such as fan pressure ratio ($\rm{PR}_{\rm{fan,TOC}}$), low pressure compressor pressure ratio ($\rm{PR}_{\rm{LPC,TOC}}$), overall pressure ratio at CRZ ($\rm{OPR}_{\rm{TOC}}$).
Other design variables are temperature controls such as burner temperature at RTO ($T_{4,\rm{RTO}}$) and the TOC-RTO burner temperature ratio ($T_{4,\rm{ratio}}$).
The nozzle velocity ratio at CRZ ($V_{\rm{ratio,CRZ}}$) is also added as a design variable to control the bypass ratio and mass flow rate through the engine.
Finally, the water recovery fraction at CRZ ($X_{\rm{H_2O,CRZ}}$) is added as a design variable in order to take advantage of water recirculation.
% [x] TODO: AL-PA, optimizing for fuel burn subject to net thrust constraint...
% [x] TODO: AL-PA, this isn't the only reason we consider just the water recovery at cruise.  It's also the longest flight segment and contributes the most to long-term performance.
We only add the water recovery fraction at CRZ as a design variable since we are optimizing $\dot{m}_{\rm{fuel,CRZ}}$.
The CRZ operating condition is also the longest flight segment and therefore contributes the most to long-term performance.
TOC-to-RTO burner temperature ratio ($T_{4,\rm{ratio}}$) is shown in Equation~\eqref{t4ratio}.
\begin{equation}
    T_{4,\rm{ratio}} = \frac{T_{4,\rm{TOC}}}{T_{4,\rm{RTO}}}
    \label{t4ratio}
\end{equation}

@@

The net thrust for the SLS and CRZ conditions are set using \emph{balance} components.
Net thrust for SLS is constrained to be about 1.2335 times the net thrust at RTO.
Similarly, the net thrust at CRZ is constrained to be 0.9 times the net thrust at TOC.
The bypass ratio (BPR) at TOC is set by the \emph{balance} component such that the CRZ nozzle velocity ratio meets a given target value.
The nozzle velocity ratio is defined as,
\begin{equation}
    V_{\rm{ratio}} = \frac{V_{\rm{core,ideal}}C_{v,\rm{core}}}{V_{\rm{bypass,ideal}}C_{v,\rm{bypass}}}
    \label{eq:vratio}
\end{equation}

\noindent
where $V_{ideal}$ is the exit velocity of a nozzle and $C_v$ is the velocity coefficient which accounts for non-ideal effects.
This adds a constraint to ensure that the core exit velocity is higher than the bypass exit velocity.
The TOC engine mass flow rate is set by the \emph{balance} component using the combustor temperature at RTO, $T_{4,\rm{RTO}}$.

% [x] TODO: AL-PA, Combine these into a single sentence.
The fuel properties used in the engine model are given in~\ref{fuel_props}.

@@

% [x] TODO: AL-PA, Define the T4 ratio with an equation and reference the equation in the table
% [x] TODO: AL-PA, I thought we were only doing recovery fractions at CRZ??
% [x] TODO: AL-PA, Make sure you reference the equations for the DVs/cons in the descriptions so the reader can connect them.
% [x] TODO: PA-PA, explain variable ranges
\begin{table}[hbt!]
    \centering
    \caption{
        Multipoint optimization problem definition.
        The objective function is the thrust specific energy consumption at the cruise condition.
        The constraints are net thrust and engine diameter constraint at the design point, TOC.
    }
    \small
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{r l l r l}
        \toprule
                        & Variable/Function              & Description                                                              & Units           & Quantity \\
        \hline
        minimize        & $\dot{m}_{\rm{fuel,CRZ}} $     & Thrust-specific energy consumption at CRZ (Equation~\eqref{eq:tsec})     & \si{lbm/hr/lbf} & 1        \\
                        &                                &                                                                          &                 &          \\
        with respect to & $X_{\rm{H_2O,CRZ}}$            & Water recovery fraction at CRZ                                           & -               & 1        \\
                        & $\rm{PR}_{\rm{fan,TOC}}$       & TOC fan pressure ratio                                                   & -               & 1        \\
                        & $\rm{PR}_{\rm{LPC,TOC}}$       & TOC low-pressure compressor pressure ratio                               & -               & 1        \\
                        & $\rm{OPR}_{\rm{TOC}}$          & TOC overall pressure ratio                                               & -               & 1        \\
                        & $T_{4,\rm{RTO}}$               & RTO combustor temperature                                                & \si{\degree R}  & 1        \\
                        & $T_{4,\rm{ratio}}$             & TOC-to-RTO temperature ratio (Equation~\eqref{t4ratio})                  & -               & 1        \\
                        & $V_{\rm{ratio,CRZ}}$           & Core-to-bypass nozzle velocity ratio at CRZ (Equation~\eqref{eq:vratio}) & -               & 1        \\
        \cline{3-5}
                        &                                & Total                                                                    &                 & 7        \\
                        &                                &                                                                          &                 &          \\
        subject to      & $F_{\rm{net,TOC}} \geq 5800.0$ & Target net thrust at TOC                                                 & \si{lbf}        & 1        \\
                        & $D_{\rm{fan}} \leq 100$        & Maximum Fan Diameter                                                     & \si{inch^2}     & 1        \\
        \cline{3-5}
                        &                                & Total                                                                    &                 & 2        \\
        \bottomrule
    \end{tabular}
    \label{tab:opt_problem}
\end{table}

@@

During initial optimization studies of this model, it was found that when $T_{4,\rm{RTO}}$ and $X_{\rm{H_2O,CRZ}}$ are design variables at the same time, they had opposite effects on the design.
The optimizer increases $X_{\rm{H_2O,CRZ}}$ which allows for a lower optimal $T_{4,\rm{RTO}}$.
This is likely due to the interdependency between engine mass flow rate and $T_{4,\rm{RTO}}$.
As the recovery water fraction increases less mass flow rate is required through the engine which lowers the required $T_{4,\rm{RTO}}$ for a required level of thrust.
It was also observed that $X_{\rm{H_2O,CRZ}}$ is increased until the model breaks due to the CEA solver not being able resolve the increased \ce{H2O} molecules.
Since $T_{4,\rm{RTO}}$ would be pushed down to its lower limit, it was set as the optimal value from the optimizations with no water recovery.
This allowed the optimizer to find the maximum $X_{\rm{H_2O,CRZ}}$ that would allow the most water to be recirculated.
The upper bound on $X_{\rm{H_2O,CRZ}}$ was found to be around 30\% for Jet-A and 17\% for H2.

@@

This is most likely due to the properties of \ce{H2} compared to Jet-A.

>>>>

% ==================================================
%	Abstract
% ==================================================
\begin{abstract}
    Advances in commercial propulsion technology led to the development of efficient high bypass ratio turbofan engines with larger overall pressure ratios and internal temperatures.
    Current trends suggest that geared ultra high bypass ratio turbofans are the next generation of commercial propulsion systems.
    Furthermore, the emphasis on decreasing emissions has driven the exploration of hydrogen-powered aircraft, adding to the already challenging design space.
    Carrying and burning hydrogen introduces complexity and weight penalties that we must offset using the fuel's thermodynamic and chemical properties.
    In this study, we create a closed-loop water recirculation system with a zero-dimensional thermodynamic model and compare the benefits between Jet-A and hydrogen fuels.
    We perform a gradient-based optimization parameter sweep to explore the trade-offs between performance and emissions using both fuels with water recirculation.
    The results quantify the design space for next-generation propulsion concepts that can take advantage of hydrogen fuel's advantageous thermodynamic properties to reduce emissions and improve performance.
\end{abstract}

\section{Introduction}
% Message 1: Motivation to incorporate low-emission fuels and techniques
The effects of climate change are pushing the aviation industry towards hydrogen-fueled propulsion systems as a solution to reduce emissions.
N+3 technology estimates for engines that burn hydrocarbon fuels suggest that higher efficiencies can be achieved by designing ultra high bypass ratio (UHBR) turbofans with small cores and high overall pressure ratios (OPR).
Higher OPR and smaller cores challenge the limits of compressor and turbine design, placing an upper bound on potential performance and emissions improvements.
Switching to hydrogen as the primary fuel source reduces carbon dioxide emissions immediately, but adds complexity and weight that can offset the benefits~\cite{Adler2023}.
However, hydrogen is a versatile fuel with advantageous molecular and thermodynamic properties that can be exploited to increase the performance and reduce emissions.
We introduce a closed-loop water recirculation model that demonstrates the possible efficiency gain when hydrogen is used for purposes other than combustion.
% [x] TODO: PA-PA, add section about MAUD and cite.

% Message 2: Background and references to support using H2 and water injection in HBTF engines
% [x] TODO: AL-PA, Need to find a journal/conference paper to cite for the following statement.  Maybe find one or more papers other than just the NASA report?
Water injection is the process of introducing water upstream of the combustor as finely atomized droplets.
NASA, Boeing, and Rolls-Royce studied this concept and suggested that this technique reduces the NOx emissions as much as 47 percent~\cite{Daggett2010}.
Additionally, water injection improves fuel efficiency and thrust output with lower combustion temperatures that can improve the lifetime of turbine blades and reduce noise~\cite{Daggett2010}.
% [x] TODO: AL-PA, Should try to find a citation for this if one exists.
Traditional propulsion systems that burn hydrocarbon fuels would require external demineralized water storage on the aircraft for water injection~\cite{Mourouzidis2015}.
The added weight of tanks, pumping, and ducting makes this concept infeasible for a conventional aircraft over the entire flight.
The main product of hydrogen combustion is water vapor that we can recover from the exhaust stream~\cite{Strom2002}.
Condensing water vapor from the exhaust stream of hydrogen combustion and recirculating it eliminates any additional storage requirements.
This allows for the theoretical design of a closed loop water feedback system inside the propulsion cycle.
Pratt and Whitney are actively researching this technology to improve the feasibility of hydrogen-powered propulsion~\cite{arpa-e_2021}.
They suggest that water vapor can be recovered by utilizing the thermal properties of hydrogen to condense some of the water in the turbofan exhaust.

@@

% Message 3: Introduce the extension of the HBTF and propose novel contributions
In this work, we analyze the thermodynamic benefits of a closed-loop water vapor recovery and water injection system in a high-bypass turbofan engine.
We develop pyCycle components for water injection and vapor recovery to quantify the benefit of a closed loop recirculation system.
We use gradient-based optimization to minimize fuel burn subject to performance requirements using both Jet-A and hydrogen at a range of flight conditions.
The optimized results show the trade-off between complexity, performance, and efficiency for Jet-A and hydrogen fuels.

@@

% Engine Architecture: Describe the flow path of the engine and establish the mechanical coupling.
The UHB turbofan model is the NASA advanced technology "N+3" engine~\cite{Jones2017a}.
The N+3 reference cycle is a UHB ratio geared turbofan that could be available in the 2030 to 2040 time frame.
The flow path consists of an inlet that directs ambient air through a fan, followed by a duct that splits the flow into a core and a bypass stream, each ending in a core and bypass nozzle, respectively.
The low pressure system is split into two mechanical subsystems.
First, the fan is connected to the gearbox that reduces the shaft speed to decrease fan tip speeds.
Second, the gearbox attaches to the low-pressure shaft that connects to the low-pressure compressor (LPC) and low-pressure turbine (LPT).
The high pressure compressor (HPC) is connected to the high pressure turbine (HPT) by the high pressure shaft.
% [x] TODO: AL-PA, I think we need to reference the engine diagram figure here so the reader can visualize all this text.
Figure~\ref{fig:n3_cycle} shows a standard high bypass turbofan engine flow connections.
% [x] TODO: AL-PA, potentially move the commented line below about CEA into the subsection discussing the propulsion model.  I think it's too much detail for the setup of this section.
We introduce the closed-loop water recovery system as a feedback cycle that transports water from the exhaust to a point upstream of the compressors.
The recovery system injects vaporized water into the core stream that reduces the combustion temperature due to heat absorption.
The vapor recovery component is placed directly before the core nozzle to extract water from the exhaust and recycles it back to the injector.
Figure~\ref{fig:n3_clvr} shows the water vapor recovery loop implemented in a turbofan engine.
% [x] TODO: AL-PA, Again, probably should reference a figure to give context to the water recovery layout.
% NOTE: AL-PA, don't discredit your work before you even explain the model and methods.  We can make this point in the results or conclusion.

@@

% [x] TODO: AL-PA, Remove the black box with the component descriptions from the bottom left of the figure.
% [x] TODO: AL-PA, Add the description of the figure elements to the caption.
% [x] TODO: AL-PA, Consider changing the caption to "Simplified layout of the N+3 engine cycle, adapted from~\citet{Hendricks2019}"
% [x] TODO: AL-PA, Remove the flight conditions component and add the bypass duct component below the splitter. I think the bypass is an important enough aspect of the engine to include.
\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_cycle.pdf}
    \caption{
        Simplified layout of the N+3 engine cycle, adapted from~\citet{Hendricks2019}.
        Black arrows are flow connections, red lines are mechanical connections, blue boxes are cycle elements, red boxes are shaft elements.
        Flow stations are identified with bubble text.
    }
    \label{fig:N3_original}
\end{figure}

@@

\begin{figure}[hbt!]
    \centering
    \includegraphics[width=1.0\textwidth]{N3_CLVR_cycle.pdf}
    \caption{
        Simplified layout of the N+3 engine cycle with the closed-loop vapor recovery.
        Black arrows are flow connections, blue boxes are cycle elements.
        Flow stations are identified with bubble text.
    }
    \label{fig:n3_clvr}
\end{figure}

@@

The ``on-design'' point is top-of-climb (TOC) with ``off-design'' conditions at rolling takeoff (RTO), sea-level static (SLS), and cruise (CRZ).
To account for each of these operating conditions, the N+3 model uses a technique called Multipoint Design Point (MDP) modeling to converge the model to an engine design.
Table~\ref{tab:flight_conds} shows the different altitudes and mach numbers for each flight condition.
We consider these points because they either limit the performance at cruise or present critical design considerations for the propulsion system.
% [x] TODO: AL-PA, Add a small table to show the mach and altitude of each flight condition
\begin{table}[hbt!]
    \centering
    \caption{Altitude and mach numbers at each of the flight conditions considered in the multipoint formulation.
        The humidity ratios were computed from humidity data~\cite{Kalnay1996}.
    }
    \begin{tabular}{l r r r r l}
        \hline
        Parameter      & TOC     & RTO   & SLS   & CRZ     & Units      \\
        \hline
        Altitude       & 35000   & 0     & 0     & 35000   & \si{ft}    \\
        Mach           & 0.8     & 0.25  & 0.0   & 0.8     & -          \\
        Humidity Ratio & 0.00017 & 0.009 & 0.009 & 0.00017 & \si{kg/kg} \\
        \hline
    \end{tabular}
    \label{tab:flight_conds}
\end{table}
% [x] TODO: AL-PA, You should explain why each design point is important in a separate sentence for each.
%   - For example, the SLS point is important for making sure the turbo machinery and flow paths are designed to meet a static thrust target
%   - The RTO point is limiting for temperature, etc...
% [x] TODO: AL-PA, Consider re-writing this section using the comments from the previous TODO.  Be more concise and direct with the language here.
% [x] TODO: AL-PA, This is a big run-on sentence.  Consider breaking it up and being more specific about the requirements at each design point.  See previous comments.
The design rules at SLS ensure that the turbo machinery and flow paths meet the static thrust target.
Rolling takeoff limitations constrain the upper limit of the combustor temperature and subsequently the cooling requirements of the turbines.
The cooling mass flow rates ($\dot{m}_{\rm{cool},k}$) are passed from the RTO point to all other flight conditions creating a cyclic connection.
We add a fuel burn objective at cruise to optimize the engine performance during the longest flight segment.
We size the water recovery system at the CRZ condition because it has the greatest impact on fuel burn over the flight envelope.
Similarly to $\dot{m}_{\rm{cool},k}$, the water recovery mass flow rates ($\dot{m}_{\rm{water},k}$) are connected back to the other operating points.
The interconnections between flight conditions create a complex nonlinear loop, shown in Figure~\ref{fig:N3_xdsm_full}.
The converged model represents a feasible multipoint engine that accounts for design considerations and conservation laws at all four operating conditions.
% [x] TODO: AL-PA, I removed a lot of stuff here to simplify this section.
%                  - Ultimately I think we need to reference the pyCycle paper to prevent duplicating a bunch of math/text
%                  - If you are okay with this check off this TODO.  If not feel free to add a bit more and then check it off.
We direct the reader to the N+3 multipoint model definition in the paper by \citet{Hendricks2019} for the detailed description of the balance relationships.

@@

Typical gas-turbine analysis measures the efficiency using the thrust-specific fuel consumption (TSFC).
% For measuring the efficiency of Jet engines, thrust-specific fuel consumption (TSFC) is used in cycle analysis since it represents how much fuel is burned at a given thrust level.
However, this metric is not useful in the comparison of efficiency between Jet-A and hydrogen fueled cycle models because of the differences in fuel density and energy capacity.
% However, when comparing Jet-A and hydrogen fuels this is not such a good metric since a given mass flow rate of hydrogen has an energy content almost 3 times that of Jet-A.
A better metric for comparing Jet-A versus hydrogen is thrust-specific energy consumption (TSEC)~\cite{Adler2023}.
TSEC is TSFC multiplied by the lower heating value (LHV) of the fuel shown in Equation~\eqref{eq:tsec}.

@@

For this work we are concerned with minimizing fuel burn of the engine subject to known thrust constraints.
% [x] TODO: AL-PA, TSEC is not the objective.  Fuel burn is the objective.  TSEC is the efficiency metric of interest.
TSEC will be used as a efficiency metric at the cruise condition to compare the two fuels in the results.
% [x] TODO: AL-PA, explain the important aspects of the XDSM in writing before referencing the figure.
The fuel properties used in the engine model are given in~\ref{fuel_props}.

Large multidisciplinary optimization problems are commonly represented with extended design structure matrix (XDSM) diagrams.
XDSM diagrams are used to visualize the complex implicit and explicit relationships between functions, the solver, and optimizer.
An XDSM diagram of the optimization problem with the multipoint formulation is shown in Figure~\ref{fig:N3_opt_xdsm}.

@@

% [x] TODO: AL-PA, I would break this up into two or more sentences.
The design variables for this problem are pressure controls such as fan pressure ratio ($\rm{PR}_{\rm{fan,TOC}}$), low pressure compressor pressure ratio ($\rm{PR}_{\rm{LPC,TOC}}$), overall pressure ratio at CRZ ($\rm{OPR}_{\rm{TOC}}$).
Other design variables are temperature controls such as burner temperature at RTO ($T_{4,\rm{RTO}}$) and the TOC-RTO burner temperature ratio ($T_{4,\rm{ratio}}$).
TOC-to-RTO burner temperature ratio ($T_{4,\rm{ratio}}$) is shown in Equation~\eqref{t4ratio}.

\begin{equation}
    T_{4,\rm{ratio}} = \frac{T_{4,\rm{TOC}}}{T_{4,\rm{RTO}}}
    \label{t4ratio}
\end{equation}

\noindent
The nozzle velocity ratio at CRZ ($V_{\rm{ratio,CRZ}}$) is also added as a design variable to control the bypass ratio and mass flow rate through the engine.
The nozzle velocity ratio is defined as,

\begin{equation}
    V_{\rm{ratio}} = \frac{V_{\rm{core,ideal}}C_{v,\rm{core}}}{V_{\rm{bypass,ideal}}C_{v,\rm{bypass}}}
    \label{eq:vratio}
\end{equation}

\noindent
where $V_{ideal}$ is the exit velocity of a nozzle and $C_v$ is the velocity coefficient which accounts for non-ideal effects.
Finally, the water recovery fraction at CRZ ($X_{\rm{H_2O,CRZ}}$) is added as a design variable in order to take advantage of water recirculation.
% [x] TODO: AL-PA, optimizing for fuel burn subject to net thrust constraint...
% [x] TODO: AL-PA, this isn't the only reason we consider just the water recovery at cruise.  It's also the longest flight segment and contributes the most to long-term performance.
We only add the water recovery fraction at CRZ as a design variable since we are optimizing $\dot{m}_{\rm{fuel,CRZ}}$.
The CRZ operating condition is also the longest flight segment and therefore contributes the most to long-term performance.

@@

The net thrust for the SLS and CRZ conditions are set using \emph{balance} components.
Net thrust for SLS is constrained to be 1.2335 times the net thrust at RTO which itself is constrained to 22800 \si{lbf}.
Similarly, the net thrust at CRZ is constrained to be 0.9 times the net thrust at TOC.
The bypass ratio (BPR) at TOC is set by the \emph{balance} component such that the CRZ nozzle velocity ratio meets a given target value.
The CRZ velocity ratio constraint ensures that the core exit velocity is higher than the bypass exit velocity.
The TOC engine mass flow rate is set by the \emph{balance} component using the combustor temperature at RTO, $T_{4,\rm{RTO}}$.
% [x] TODO: AL-PA, Combine these into a single sentence.

@@

% [x] TODO: AL-PA, Define the T4 ratio with an equation and reference the equation in the table
% [x] TODO: AL-PA, I thought we were only doing recovery fractions at CRZ??
% [x] TODO: AL-PA, Make sure you reference the equations for the DVs/cons in the descriptions so the reader can connect them.
% [x] TODO: PA-PA, explain variable ranges
\begin{table}[hbt!]
    \centering
    \caption{
        Multipoint optimization problem definition.
        The objective function is the thrust specific energy consumption at the cruise condition.
        The constraints are net thrust and engine diameter constraint at the design point, TOC.
    }
    \small
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{r l l r l}
        \toprule
                        & Variable/Function              & Description                                                              & Units          & Quantity \\
        \hline
        minimize        & $\dot{m}_{\rm{fuel,CRZ}} $     & Fuel flow rate at CRZ                                                    & \si{lbm/s}     & 1        \\
                        &                                &                                                                          &                &          \\
        with respect to & $X_{\rm{H_2O,CRZ}}$            & Water recovery mass fraction at CRZ                                      & -              & 1        \\
                        & $\rm{PR}_{\rm{fan,TOC}}$       & TOC fan pressure ratio                                                   & -              & 1        \\
                        & $\rm{PR}_{\rm{LPC,TOC}}$       & TOC low-pressure compressor pressure ratio                               & -              & 1        \\
                        & $\rm{OPR}_{\rm{TOC}}$          & TOC overall pressure ratio                                               & -              & 1        \\
                        & $T_{4,\rm{RTO}}$               & RTO combustor temperature                                                & \si{\degree R} & 1        \\
                        & $T_{4,\rm{ratio}}$             & TOC-to-RTO temperature ratio (Equation~\eqref{t4ratio})                  & -              & 1        \\
                        & $V_{\rm{ratio,CRZ}}$           & Core-to-bypass nozzle velocity ratio at CRZ (Equation~\eqref{eq:vratio}) & -              & 1        \\
        \cline{3-5}
                        &                                & Total                                                                    &                & 7        \\
                        &                                &                                                                          &                &          \\
        subject to      & $F_{\rm{net,TOC}} \geq 5800.0$ & Target net thrust at TOC                                                 & \si{lbf}       & 1        \\
                        & $D_{\rm{fan}} \leq 100$        & Maximum Fan Diameter                                                     & \si{inch^2}    & 1        \\
        \cline{3-5}
                        &                                & Total                                                                    &                & 2        \\
        \bottomrule
    \end{tabular}
    \label{tab:opt_problem}
\end{table}

@@

During initial optimization studies of this model, it was found that when $T_{4,\rm{RTO}}$ and $X_{\rm{H_2O,CRZ}}$ are design variables at the same time they had opposite effects on the design.
The optimizer increases $X_{\rm{H_2O,CRZ}}$ which allows for a lower optimal $T_{4,\rm{RTO}}$.
This is likely due to the interdependency between engine mass flow rate and $T_{4,\rm{RTO}}$.
As the recovery water fraction increases less mass flow rate is required through the engine which lowers the required $T_{4,\rm{RTO}}$ for a required level of thrust.
It was also observed that $X_{\rm{H_2O,CRZ}}$ is increased until the model breaks due to the CEA solver not being able resolve the increased \ce{H2O} molecules that are injected.
Since $T_{4,\rm{RTO}}$ would be pushed down to its lower limit, it was set as the optimal value from the optimizations with no water recovery.
This allowed the optimizer to find the maximum $X_{\rm{H_2O,CRZ}}$ that would allow the most water to be recirculated.
The upper bound on $X_{\rm{H_2O,CRZ}}$ was found to be around 30\% for Jet-A and 17\% for \ce{H2}.

@@

This is most likely due to the combustion properties of \ce{H2} compared to Jet-A.
